TacticToe: Learning the Game of Tactics

### 2018-01-29 tactictoe_jv.tex
search: first final version

\begin{definition}\label{def:desc}(Children, Descendant)\\
We now define what are the children and the descendants of a node $a$ in a 
search tree $\mathfrak{T}$. The children of a relative to a goal $g$ in $G(a)$ 
is the set of nodes produced by tactics applied to $g$:
  \[Children(a,g) = \lbrace A(g,t)\in \mathbb{A}\ |\ t \in T(g) \rbrace \]

From that, we define the children of a node $a$ by:
  \[Children(a) = \bigcup_{g \in G(a)} Children(a,g) \]

And descendants of $a$ by transitive closure:
\begin{align*}
Descendant^{0}(a) &=_{def} \lbrace a \rbrace \\ 
Descendant^{n+1}(a) &=_{def} \bigcup_{a' \in Descendant^{n}(a)} 
Children(a') \\
Descendant(a) &=_{def} \bigcup_{n \in \mathbb{N}} Descendant^n(a)\\
\end{align*}

Starting from the bottom parent node, we can construct children of a goal of 
this node by choosing a goal $g_i^0$ on this node and choosing tactics 
$t_1,t_2,\ldots$ to 
apply to this goal. Here the result produces new nodes each containing a list 
of goals to be proven. In each node of the proof tree, a goal can either be 
solved, open or pending (see Definition~\ref{def:solved}). 
\end{definition}

\begin{definition}\label{def:solved}(Solved goal, solved nodes, open goal, 
pending goal , active node)\\
Let $\mathfrak{T}$ be a search tree. The set of solved nodes 
$\mathbb{A}^\infty$ and 
the set of solved goals $\mathbb{G}^\infty$ are defined inductively by:

@@

\todo{make definition in preliminaries of active node and descendant?}

@@

The third point of the definition is a partial attempt at preventing confluent 
branches. The general case where two branches join after $n$ steps is handled 
by a tactic cache which memorizes tactic effects on goals. This
cache allows faster re-exploration of confluent branches which remain 
separated in the search tree.

\todo{make definition in preliminaries of active node and descendant?}

@@

\paragraph{List of Goals as Output of Tactics}
A dataset of tactic outputs is compiled during orthogonalization (see 
Section~\ref{sec:evaluation}).
This set is separated into positive and negative examples.
List of goals that subsumes the result of an original tactic on a 
targeted goal are considered positive. During proof search, given a list of 
goals $l$ created by a tactic, we select a set of lists of goals that are most 
similar to $l$. The ratio of positive examples in the selection gives us a 
prior node evaluation for MCTS.

@@

\subsection{Prior tactic policy}\label{sec:policy}

\todoi{mention productive tactics}
Distances between goals produced by the tactic predictor are hard 
to translate into a prior tactic policy, so we rely solely on the order of 
tactics to estimate the probability with which each tactic should be tried.
Let $t_0,\ldots,t_n$ be the list of preselected tactics ordered by their 
distance to a goal $g$. Let $c_{policy}$ be a constant heuristic that estimates 
how likely a tactic $t_i$ is to prove the goal compared to the rest of the 
tactics $t_{i+1},\ldots,t_{500}$. We make the strong assumption that $c_policy$ 
is constant across the search. We can now calculate the prior policy by:

\[PriorPolicy(a_i) = PriorPolicy(t_i) = PriorPolicy(i) = (1 - c_{policy})^{i} * 
c_{policy}\]

Widening policy for a node $a$.

\[WideningPolicy(a) = \]

The first term of the multiplication is the probability that none of the better 
ranked tactic is selected. In fact, these probabilities are never used directly 
and only serves as a guidance to our deterministic node selection algorithm 
over many MCTS step.

\subsection{Prior node evaluation}\label{sec:evaluation}

We now concentrate on defining a reasonable evaluation function that also 
influences node selection in MCTS. Essentially, we estimate how 
interesting is a list of goals produced by a tactic. This way, we can determine 
if we should continue searching in this direction. Theoretically, an 
interesting list of goal is one 
that has the shortest proof, and we could measure the likelihood of being 
provable in a number of steps from previous proof search. However, this 
approach produces a lot of negatives\todo{CK: cite hard negative mining 
or/and deep math}. So to reduce this number, we limit 
ourselves to tactics tested during orthogonalization. We declare a list 
of goals l to be positive, noted $l \in Positive$, if it is has been 
produced by the winner of an orthogonalization competition. The set of the $r$ 
goals closest to a list of goals $l$ is noted $Neighbour(r,l)$. We can 
define the set of positives within a 
radius $r$ of $l$ to be $Positive(r,l) = \lbrace m \in Neighbour(r,l)\ |\ m \in 
Positive \rbrace$.

During our experiments, we evaluate output goals with the following function:

\begin{align*}
PriorEvaluation_r (a_i) &= \frac{card\ Positive(r,a_i)}{r}\\
\end{align*}

If a tactic is not productive then there is no output to evaluate, so the 
result is a failure and a reward of $0$ will be 
given for this node. Failure also happens if the node selected is already 
a proven goal or a descendant of a proven goal. Experimentally, it is 
preferable not 
to filter solved nodes and descendants of solved nodes before node selection 
to keep the estimations of the algorithm balanced.

@@

Through node extension steps the search tree grows and the number of paths to 
explore increases. To decide which node to extend next, we proceed through the 
node selection algorithm. Starting from the root node, we repeatedly chose the 
children relative to the open goal with the highest value.

The value of a $a_i$ with respect to a search tree is determined by:
\[Value(a_i) = CurEvaluation(a_i) + c_{exploration} * Exploration(a_i)\]

@@

The exploitation term $CurEvaluation(a_i)$ is the average of the evaluation of 
all MCTS step that went through $a_i$ (itself included) in the current search 
tree.

\[CurEvaluation(a_i) = 
  \sum_{a' \in Descendant(a_i)} \frac{PriorEvaluation(a')} {card\ 
  Descendant(a_i) 
  + card\ Failure(a_i)}\]

where $Failure(a_i)$ is the number of MCTS steps that went through $a_i$ which 
ended in a failure.

To allow the algorithm to select nodes that 
are not leafs, we also assign a value to the possibility of stopping during the 
descent of the tree. This value has an exploitation equal to 0 since it was 
never explored and is given a prior policy equal to $PriorPolicy(n)$ where $n$ 
is the 
number of 
tactics already tried on the open goal if there is one and $PriorPolicy(0)$ if 
there is 
none.

\subsection{Node extension}
Once a node is selected, we first check if it is an active part of the search 
tree (not a solved node .

If the selected node is not active, then the result of node extension is a 
failure.
Otherwise, we apply the best untested tactic on the open goal $g$ of this node  
according to the prediction order for $g$.

we apply the next tactic in the prediction order for 
the open goal of this node.

The result of the extension step is either a failure or a list of goals $l$.
From $l$, we can assess if the tactic was productive. 
If not, we also return a failure.
Otherwise, the list of goals $l$ is used to create a new node in the search 
tree. Give precise definition.

order of predicted tactics.

During the creation of node, we predict tactics for the open goals

\subsection{Backpropagation}

If node extension resulted in a failure, we increment the failure field of the 
selected node and its parents. If node extension was successful, we 
backpropagate the evaluation of the new nodes to their parents.
In both cases, we increment the number of visits of all nodes traversed.
This changes induce an update on the current evaluation and current policy for 
each node. And the process loops back to the node selection step.

@@

Such a prover already exists in \holfour and is called \metis. It is already 
recognized by the tactic selection mechanism in \tactictoe and can have its 
premises predicted dynamically when in an abstracted form. Nevertheless,
we think that the performance of \tactictoe can be boosted by giving the ATP 
\metis a special status. This arrangement consists in always predicting 
premises for \metis, giving it a slightly higher timeout than other tactics and 
always trying it first on each node. These modifications only induce a
overhead linear on the number nodes as a general purpose prover does not create
goal obligations.

The integration of an external prover is also possible and 
\eprover~\cite{eprover} is the one we experimented with .
To this end, we created a tactic which expects a number of premises as argument 
and calls \eprover on a version of the problem (premises + goal) translated to 
first order. We then extract the premises that were required in the \eprover 
proof and reconstruct it with \metis. Giving a special status to external 
prover is really necessary as external provers calls do not appear in human 
proof scripts. The setting in place is about the same as for \metis, the 
difference being an even higher timeout (5 seconds) and a larger number of 
predicted premises (128). Since calls to \eprover are
computationally, they are run in parallel and asynchronously. This 
avoids slowing down \tactictoe's search loop. The number of asynchronous calls 
to an external provers that can be executed at the same time is limited by the 
number of cores available to the user.

%can be call with frequently with low 
%timeout.
%And the strongest external provers according to past experiments .
%Where as...
%
%experiment with a small hammer tactic comprised of: a faster premise 
%selection algorithm and
%a short call (0.1 seconds + 16 premises optimized by the previous paper) to 
%the 
%internal prover \metis~\cite{metis} or a
%longer call to external prover \eprover. 
%
%
%\paragraph{Prediction of general purpose tactics}
%Since we do not always have the data to learn on which goal this general 
%tactics are successful, we assume as a first approximation that they are the 
%strongest all around. 
%
%And we try them first on each node during the proof search with the advantage 
%that if it succeeds the node is closed rapidly.
%
%This general tactics requires a list of relevant lemmas.
%
%This forced behavior could often be rediscovered from the training data for 
%the 
%internal provers \metis by using theorem abstraction. essentially a call to 
%\metis where its 
%theorem list argument has been abstracted.
%
%
%\paragraph{Application of general purpose tactics}
%The slightly higher timeout of 0.1 seconds given to \metis does not slowdown 
%the proof too much but...
%To integrate a general tactic that requires a higher time out such as 
%\eprover,

>>>>

Starting from the bottom parent node, we can construct children of a goal of 
this node by choosing a goal $g_i^0$ on this node and choosing tactics 
$t_1,t_2,\ldots$ to 
apply to this goal. Here the result produces new nodes each containing a list 
of goals to be proven. In each node of the proof tree, a goal can either be 
solved, open or pending (see Definition~\ref{def:solved}).

In the following, new concepts and properties are defined in the context of a 
search tree $\mathfrak{T}$.

\begin{definition}\label{def:solved}(Solved goal, solved nodes, open goal, 
pending goal)\\
The set of solved nodes $\mathbb{A}^\infty$ and 
the set of solved goals $\mathbb{G}^\infty$ are defined inductively by:

@@

@@

The third point of the definition is a partial attempt at preventing confluent 
branches. The general case where two branches join after $n$ steps is handled 
by a tactic cache which memorizes tactic effects on goals. This
cache allows faster re-exploration of confluent branches which remain 
separated in the search tree.
\end{definition}

\begin{definition}\label{def:desc}(Children, descendant, active node)\\
We now define what are the children and the descendants of a node $a$. The 
children of a relative to a goal $g$ in $G(a)$ 
is the set of nodes produced by tactics applied to $g$:
  \[Children(a,g) = \lbrace A(g,t)\in \mathbb{A}\ |\ t \in T(g) \rbrace \]

If $g$ is the open goal of $a$, then we simply call children of a the 
children of $a$ relative to $g$. We define:\\

Extended children of $a$ by:
  \[ExtChildren(a) = \bigcup_{g \in G(a)} Children(a,g) \]

Descendants of $a$ by:
\begin{align*}
Descendant^{0}(a) &=_{def} \lbrace a \rbrace \\ 
Descendant^{n+1}(a) &=_{def} \bigcup_{a' \in Descendant^{n}(a)} 
ExtChildren(a') \\
Descendant(a) &=_{def} \bigcup_{n \in \mathbb{N}} Descendant^n(a)\\
\end{align*}

Descendants of $a$ through $g$ by:
\[Descendant(a,g) =_{def} 
  \lbrace a \rbrace \cup \bigcup_{a' \in Children(a,g)} Descendant(a')\]

An active node is by definition a node that is not a descendant of a 
solved node. Non-active nodes are not worth exploring.

@@

\paragraph{List of Goals as Output of Tactics}
A dataset of tactic outputs is compiled during orthogonalization (see 
Section~\ref{sec:evaluation}).
This set is separated into positive and negative examples.
List of goals that subsumes the result of an original tactic on a 
targeted goal are considered positive. During proof search, given a list of 
goals $l$ created by a tactic, we select a set of lists of goals that are most 
similar to $l$. The ratio of positive examples in the selection gives us a 
prior evaluation for MCTS.

@@

\subsection{Prior policy}\label{sec:policy}
Co-distances between goals produced by the tactic predictor are hard 
to translate into a prior policy, so we rely solely on the order of 
tactics to estimate the probability with which each tactic should be tried.
Let $\mathfrak{T}$ be a search tree after a number of MCTS steps.
Let $p$ be a selected node on this tree and $g$ its open goal. By the 
application of MCTS, $T(g)$ is the list of $n$ productive tactics already 
applied to $g$. We order them by their prediction score so that $T(G)= \lbrace 
t_0,\ldots,t_{n-1} \rbrace$. 
Let $c_{policy}$ be a constant heuristic that estimates how likely a tactic 
$t_i$ is to prove the goal. 
We can now calculate of a prior policy of a child $a_i$ produced by the tactic 
$t_i$ by:

\[PriorPolicy(a_i) = (1 - c_{policy})^{i} * c_{policy}\]

In order to include the possibly of trying more tactic on $g$ we define the 
widening policy on the parent $a$ for its open goal $g$ to be:

\[WideningPolicy(p) = (1 - c_{policy})^{n} * c_{policy}\]

\subsection{Prior evaluation}\label{sec:evaluation}

We now concentrate on the definition a reasonable evaluation function for the 
output of tactics. Theoretically, an interesting list of goal is one 
that has a short proof, and we could measure the likelihood of being 
provable in a number of steps from previous proof search. However, this 
approach produces a lot of negatives\todo{CK: cite hard negative mining 
or/and deep math}. So to reduce this number, we limit 
ourselves to collect output of tactics tested during orthogonalization. We 
declare a list of goals $l$ to be positive if it is has been produced by the 
winner of an orthogonalization competition. The predictor with radius $r$ 
\todo{define radius} produces a set of $r$ list of goals closest to $l$. The 
number of positive example in this set is noted $Positive(r,l)$.
During the proof search, we evaluate nodes through the list of goals they 
contain using the following evaluation function:
\begin{align*}
PriorEvaluation_r (a_i) &= \frac{Positive(r,G(a_i))}{r}\\
\end{align*}

@@

Through node extension steps the search tree grows and the number of paths to 
explore increases. To decide which node to extend next, a evolving value for 
each node is given in Definition~\ref{def:value}. The algorithm that performs 
node selection starts from the root of the search tree. X, it repeatedly choses 
the children with the highest values. Depending on the widening policy, it can 
stop at any chosen node on during its descent through the tree. We now give a 
more precise description of 
the algorithm with the following pseudo-code:

\todoi{Replace alltt environment by something prettier}
\begin{alltt}
ChosenNode = Root(Tree);
while true do
  if Children(ChosenNode) = \(\emptyset\) then break;
  (BestChild, BestValue) = CompareChildren (ChosenNode);
  if WideningPolicy (ChosenNode) \(\geq\) BestValue then break;
  ChosenNode = BestChild
end;
return ChosenNode;
\end{alltt}

\begin{definition}\label{def:value}(Value)
The value of the $i^th$ child $a_i$ of a parent $p$ is determined by:
\[Value(a_i) = CurEvaluation(a_i) + c_{exploration} * Exploration(a_i)\]

@@

The current evaluation $CurEvaluation(a_i)$ is the average evaluation of 
all descendants of $a_i$, through its open goal $g_i$, including node extension 
failures:

\[CurEvaluation(a_i) = 
  \sum_{a' \in Descendant(a_i,g_i)} \frac{PriorEvaluation(a')} {card\ 
  Descendant(a_i,g_i) + Failure(a_i)}\]

The number $Failure(a_i)$ is the amount of failures that occurred during node 
extension from descendants of $a_i$ .
\end{definition}

\subsection{Node extension}
Let $a$ be the result of node selection.
If $a$ is not an active part of the search then extending $a$ is useless and 
the algorithm reports a failure. In our experiments, filtering nodes that not 
active before node selection is 
not beneficial. It may be because it breaks some symmetry in the 
$Value$ function. 
Otherwise, it applies the best untested tactic $t$ on the open goal $g$ of this 
node according to the prediction order for $g$. If no such tactic exists or if 
$t$ is not productive, a failure is returned. 
Otherwise, a new node $c$ is created that contains the list of goals $l$ 
produced by $t$ and is added to the children of $a$.

\subsection{Backpropagation}
During backpropagation we update the statisitics of all the nodes traversed or 
created during this MCTS step:
\begin{itemize}
\item Their number of visit is incremented.
\item If node extension failed, their failure count is incremented.
\item If node extension succeeded, they inherit the evaluation of the created 
child. 
\end{itemize}

These changes updates the current evaluation and the current policy of the 
traversed nodes. 
After completing backpropagation, the process loops back to node 
selection.

@@

Such a prover already exists in \holfour and is called \metis. It is already 
recognized by the tactic selection mechanism in \tactictoe and can have its 
premises predicted dynamically when in an abstracted form. Nevertheless,
we think that the performance of \tactictoe can be boosted by giving the ATP 
\metis a special status. This arrangement consists in always predicting 
premises for \metis, giving it a slightly higher timeout and 
always trying it first on each node. These modifications only induce an
overhead linear on the number nodes as a general purpose prover does not create
goal obligations. \todoi{Define goal obligation in the introduction}

The integration of an external prover is also possible and 
\eprover~\cite{eprover} is the one we experimented with .
To this end, we created a tactic which expects a number of premises as argument 
and calls \eprover on the translated first order problem. We then extract the
premises that were required in the \eprover 
proof and reconstruct it inside \holfour with \metis. Giving a special status 
to external 
prover is essential as external provers calls do not appear in human
proof scripts. The setting in place is about the same as for \metis, the 
difference being an even higher timeout (5 seconds) and a larger number of 
predicted premises (128). Since calls to \eprover are
computationally expensive, they are run in parallel and asynchronously. This 
avoids slowing down \tactictoe's search loop. The number of asynchronous calls 
to an external provers that can be executed at the same time is limited by the 
number of cores available to the user.
